**Master analysis for skin microbiome project**
========================================================

```{r,echo=FALSE}
#Define file paths to data files
setwd("~/Documents/UTS/skin_microbiome/Samples/Westmead/data/run4/UPARSE2/")
#Show boxplots of distances between diabetic skin types and wounds, and do significance testing. Beta-diversity was calculated in Qiime on a 30K rarefied OTU table using unifrac, and the make_distance_boxplots.py command to compare distnaces between groups.
dist<-read.csv("distances_contra_adj_v_wounds2.csv", header=TRUE)
```
Generate boxplots based on weighted unifrac distances between diabetic adjacent and contrateral skin, and wound swabs and debridement
```{r, echo=FALSE, fig.width=10, fig.height=5}
boxplot(dist, las=2, main="Unifrac distances between adjacent or contralateral\n diabetic skin and wounds", ylab="Weighted unifrac distance", cex.lab=0.75, cex.main=0.75, cex.axis=0.75, ylim=c(0,0.8))
```
Test for significant differences between groups
```{r, echo=FALSE}
ks.test(dist$adj_V_ws, dist$con_V_ws)
ks.test(dist$adj_V_wd, dist$con_V_wd)
ks.test(dist$adj_V_wd, dist$adj_V_ws)
ks.test(dist$con_V_wd, dist$con_V_ws)
```
This shows that adjacent diabetic skin is (slightly) more similar to wounds than adjacent diabetic skin, and this difference is significantly different. This justifies the removal of the adjacent diabetic skin samples from the analysis.

Analysis of richness from rarefied OTU table
```{r, echo=FALSE}
#define paths to files
biom=("r_data/diabetes_all_uparse_tax_goodalign_json.biom")
map=("r_data/mapping_diabetes_all_uparse.txt")
tree=("r_data/otus_cf_aligned.tre")
#import data
library("phyloseq")
dswd <- import_qiime_sample_data(map)
biom_otu_table<- import_biom(biom, tree)
mydata <- merge_phyloseq(biom_otu_table,dswd)
colnames(tax_table(mydata)) = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")
library("ggplot2")
theme_set(theme_bw())
#Subset data to exclude positive, negative 
mydata_sub<-subset_samples(mydata, sampletype1%in%c("diabetic_skin","control_skin","wound"))
#Subset again to remove diabetic_skin_adj
mydata_sub<-subset_samples(mydata_sub, sampletype2!="diabetic_skin_adj")
#subset again to remove P3 and P4
mydata_sub<-subset_samples(mydata_sub, !(subject%in%c("P3","P4")))
#Remove samples with less than 30K reads
mydata_sub<-prune_samples(names(which(sample_sums(mydata_sub) >= 30000)), mydata_sub)
#rarefy otu table to 30K reads
rare_mydata_sub<-rarefy_even_depth(mydata_sub, sample.size=30000, trimOTUs=TRUE, rngseed=552)
```
```{r, echo=FALSE, fig.width=10, fig.height=4}
figure_1<-phyloseq::plot_richness(rare_mydata_sub, x="sampletype1", measures=c("Observed", "Chao1","Shannon"))+geom_boxplot()+xlab("Sample type")+ylab("Diversity")
figure_1
```
Significance testing on alpha diversity results
```{r, echo=FALSE}
alpha_estimates<-estimate_richness(rare_mydata_sub, split = TRUE, measures = c("Observed", "Chao1", "Shannon"))
#Sort samples in alpha_estimates by sample name
alpha_estimates_sorted<-alpha_estimates[order(rownames(alpha_estimates)),]
#sort samples in sample_data(rare_mydata_sub) by sample name
sorted_sampledata<-sample_data(rare_mydata_sub)[order(rownames(sample_data(rare_mydata_sub))),]
#Create a vector of which group each sample belongs to from the sampletype1 column and add to alpha estimates file
group<-sorted_sampledata$sampletype1
alpha_estimates_sorted<-cbind(alpha_estimates_sorted, group)
rm(group)
#Sort sorted_alpha_estimates by group
alpha_estimates_sorted<-alpha_estimates_sorted[order(alpha_estimates_sorted$group),]
#Get median, min and max "Observed" values for the three groups under "group"
library(plyr)
median_otus<-ddply(alpha_estimates_sorted, .(group), summarize, median_otus=median(Observed), max=max(Observed), min=min(Observed))
median_otus
#Code from Steve Bush to assess statistical difference in diversity between groups.
library(MASS)
fullmodel<-glm.nb(Observed ~factor(group),data=alpha_estimates_sorted)
reducedmodel<- glm.nb (Observed ~1,data=alpha_estimates_sorted)
library(lmtest)
lrtest(fullmodel, reducedmodel)
#Wilcox rank sum test
wilcox.test(Observed ~ group, alpha_estimates_sorted, subset=group%in%c("control_skin", "diabetic_skin"))
wilcox.test(Observed ~ group, alpha_estimates_sorted, subset=group%in%c("wound", "diabetic_skin"))
wilcox.test(Observed ~ group, alpha_estimates_sorted, subset=group%in%c("wound", "control_skin"))
#Now stats for Chao1 and Shannon
fullmodel<-glm.nb(Chao1 ~factor(group),data=alpha_estimates_sorted)
reducedmodel<- glm.nb (Chao1 ~1,data=alpha_estimates_sorted)
lrtest(fullmodel, reducedmodel)
wilcox.test(Chao1 ~ group, alpha_estimates_sorted, subset=group%in%c("control_skin", "diabetic_skin"))
wilcox.test(Chao1 ~ group, alpha_estimates_sorted, subset=group%in%c("wound", "diabetic_skin"))
wilcox.test(Chao1 ~ group, alpha_estimates_sorted, subset=group%in%c("wound", "control_skin"))
fullmodel<-glm.nb(Shannon ~factor(group),data=alpha_estimates_sorted)
reducedmodel<- glm.nb (Shannon ~1,data=alpha_estimates_sorted)
lrtest(fullmodel, reducedmodel)
wilcox.test(Shannon ~ group, alpha_estimates_sorted, subset=group%in%c("control_skin", "diabetic_skin"))
wilcox.test(Shannon ~ group, alpha_estimates_sorted, subset=group%in%c("control_skin", "diabetic_skin"), correct=FALSE)
wilcox.test(Shannon ~ group, alpha_estimates_sorted, subset=group%in%c("wound", "diabetic_skin"))
wilcox.test(Shannon ~ group, alpha_estimates_sorted, subset=group%in%c("wound", "control_skin"))
```
Plots of top 10 OTUs by average abundance per group.
```{r, echo=FALSE}
#Using apply to get average abundances of OTUs per group.  First create vectors of names of each group of samples
control_skin_samples<-rownames(sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"sampletype1"]=="control_skin"])
diabetic_skin_samples<-rownames(sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"sampletype1"]=="diabetic_skin"])
wound_swab_samples<-rownames(sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"sampletype2"]=="wound_swab"])
wound_deb_samples<-rownames(sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"sampletype2"]=="wound_deb"])
#Now apply the mean function only to columns of the OTU table that correspond to a sample from each of the vectors of groups sample names. i.e. gets mean abundance for each OTU for each of the 4 sample groups defined above.
mean_abund_control_skin_otus<-as.matrix(apply(otu_table(rare_mydata_sub)[,control_skin_samples],1, FUN=mean))
stdev_abund_control_skin_otus<-as.matrix(apply(otu_table(rare_mydata_sub)[,control_skin_samples],1, FUN=sd))
mean_abund_diabetic_skin_otus<-as.matrix(apply(otu_table(rare_mydata_sub)[,diabetic_skin_samples],1, FUN=mean))
stdev_abund_diabetic_skin_otus<-as.matrix(apply(otu_table(rare_mydata_sub)[,diabetic_skin_samples],1, FUN=sd))
mean_abund_wound_swab_otus<-as.matrix(apply(otu_table(rare_mydata_sub)[,wound_swab_samples],1, FUN=mean))
mean_abund_wound_deb_otus<-as.matrix(apply(otu_table(rare_mydata_sub)[,wound_deb_samples],1, FUN=mean))
#Now bind those results into a new table and define row (OTU) and column (sample) names.
average_abund_all<-as.matrix(cbind(mean_abund_control_skin_otus,mean_abund_diabetic_skin_otus,mean_abund_wound_swab_otus,mean_abund_wound_deb_otus))
rownames(average_abund_all)<-rownames(otu_table(rare_mydata_sub))
colnames(average_abund_all)<-c("control skin","diabetic skin","wound swabs","wound debridement")

#Still need to figure our best colour scale and labels - will do this later.
```
Plots of top 10 OTUs by average abundance per group in skin only.
```{r, echo=FALSE}
#Bind results from above into a new table and define row (OTU) and column (sample) names.
average_abund_skin<-as.matrix(cbind(mean_abund_control_skin_otus,mean_abund_diabetic_skin_otus))
rownames(average_abund_skin)<-rownames(otu_table(rare_mydata_sub))
colnames(average_abund_skin)<-c("control skin","diabetic skin")
#Now sort table by the 20 most abundant across all sample types, after converting counts to propotions.
average_abund_skin<-((average_abund_skin/30000)*100)
average_abund_skin<-average_abund_skin[order(-rowSums(average_abund_skin)),]
#Find top 10
top10skin<-average_abund_skin[1:10,]
top10skin<-data.frame(top10skin)
top10skin<-cbind(top10skin,tax_table(rare_mydata_sub)[rownames(top10skin),])
top10skin<-top10skin[,c(1,2,7,8,9)]
top10skin$OTU<-paste(top10skin$Family,top10skin$Genus, top10skin$Species, rownames(top10skin))
#Write file for making nice graphics in another program if necessary
write.table(top10skin, file="top10skin2.txt", sep="\t")
top10skin<-top10skin[,c(1, 2,6)]
#Now melt data to the right format for ggplot
top10skin<-melt(top10skin)
#Now plot a stacked barplot
figure2<-ggplot(top10skin, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+ labs(title="Top 10 OTUs in control and diabetic skin", x="", y="Relative abundance (%)")+ theme(legend.text=element_text(size=8), legend.title=element_text(size=10),legend.key.size = unit(0.5, "cm"))+scale_fill_brewer(palette="Paired")
figure2
rare_skin<-subset_samples(rare_mydata_sub, sampletype1%in%c("control_skin","diabetic_skin"))
OTU_13<-t(data.frame(otu_table(rare_skin)["OTU_13",]))
OTU_13<-OTU_13[order(rownames(OTU_13)),,drop=FALSE]
subject<-sample_data(rare_skin)[order(rownames(sample_data(rare_skin))),"subject"]
OTU_13<-cbind(OTU_13,subject)
OTU_13$"OTU_13"<-(OTU_13$"OTU_13"/30000)*100
ggplot(OTU_13, aes(x=subject, y=OTU_13)) + 
  geom_boxplot()+ylim(0,5)+labs(x = "Subject", y="Relative abundance (%)")+
  ggtitle("Kocuria palustris relative abundance by subject")
OTU_5109<-t(data.frame(otu_table(rare_skin)["OTU_5109",]))
OTU_5109<-OTU_5109[order(rownames(OTU_5109)),,drop=FALSE]
OTU_5109<-cbind(OTU_5109,subject)
OTU_5109$"OTU_5109"<-(OTU_5109$"OTU_5109"/30000)*100
ggplot(OTU_5109, aes(x=subject, y=OTU_5109)) + 
  geom_boxplot()+labs(x = "Subject", y="Relative abundance (%)")+
  ggtitle("Staphylococcus pettenkoferi relative abundance by subject")
```
Skin top 10 over time per subject
```{r}
#make otu table only for P1 wound swabs, and sort by rowsums.  Then convert counts to %, order columns by name, add taxonomy, bind to a single OTU columns with the numerical OTU identifier, subset to top ten and plot over time using ggplot.
  P1sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="P1"]
  P1_skin<-rownames(P1sd)[P1sd[,"sampletype2"]=="diabetic_skin_contra"]
  P1_skin_abund<-data.frame(otu_table(rare_mydata_sub)[,P1_skin])
  P1_skin_abund<-P1_skin_abund[order(-rowSums(P1_skin_abund)),]
  P1_skin_abund<-(P1_skin_abund/30000)*100
  P1_skin_abund<-P1_skin_abund[,order(names(P1_skin_abund))]
  P1_skin_abund<-cbind(P1_skin_abund,tax_table(rare_mydata_sub)[rownames(P1_skin_abund),c("Family","Genus","Species")])
  P1_skin_abund$OTU<-paste(P1_skin_abund$Family,P1_skin_abund$Genus,P1_skin_abund$Species, rownames(P1_skin_abund))
  P1_skin_abund<-P1_skin_abund[,c(1,2,3,4,5,6,10)]
  P1_skin_top10<-P1_skin_abund[1:10,]
  colnames(P1_skin_top10)<-c("0","1","2","3","4","5","OTU")
  P1_skin_top10<-melt(P1_skin_top10)
  #Now plot a stacked barplot
  P1_skin_plot<-ggplot(P1_skin_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Patient 1")+ labs(x = "Time point", y="Relative abundance")
  P1_skin_plot
  #make otu table only for P2 wound swabs, and sort by rowsums.  Then convert counts to %, order columns by name, add taxonomy, bind to a single OTU columns with the numerical OTU identifier, subset to top ten and plot over time using ggplot.
  P2sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="P2"]
  P2_skin<-rownames(P2sd)[P2sd[,"sampletype2"]=="diabetic_skin_contra"]
  P2_skin_abund<-data.frame(otu_table(rare_mydata_sub)[,P2_skin])
  P2_skin_abund<-P2_skin_abund[order(-rowSums(P2_skin_abund)),]
  P2_skin_abund<-(P2_skin_abund/30000)*100
  P2_skin_abund<-P2_skin_abund[,order(names(P2_skin_abund))]
  P2_skin_abund<-cbind(P2_skin_abund,tax_table(rare_mydata_sub)[rownames(P2_skin_abund),c("Family","Genus","Species")])
  P2_skin_abund$OTU<-paste(P2_skin_abund$Family,P2_skin_abund$Genus,P2_skin_abund$Species, rownames(P2_skin_abund))
  P2_skin_abund<-P2_skin_abund[,c(1,2,3,7)]
  P2_skin_top10<-P2_skin_abund[1:10,]
  colnames(P2_skin_top10)<-c("0","4","5","OTU")
  P2_skin_top10<-melt(P2_skin_top10)
  #Now plot a stacked barplot
  P2_skin_plot<-ggplot(P2_skin_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Patient 2")+ labs(x = "Time point", y="Relative abundance")
  P2_skin_plot
  #make otu table only for P5 wound swabs, and sort by rowsums.  Then convert counts to %, order columns by name, add taxonomy, bind to a single OTU columns with the numerical OTU identifier, subset to top ten and plot over time using ggplot.
  P5sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="P5"]
  P5_skin<-rownames(P5sd)[P5sd[,"sampletype2"]=="diabetic_skin_contra"]
  P5_skin_abund<-data.frame(otu_table(rare_mydata_sub)[,P5_skin])
  P5_skin_abund<-P5_skin_abund[order(-rowSums(P5_skin_abund)),]
  P5_skin_abund<-(P5_skin_abund/30000)*100
  P5_skin_abund<-P5_skin_abund[,order(names(P5_skin_abund))]
  P5_skin_abund<-cbind(P5_skin_abund,tax_table(rare_mydata_sub)[rownames(P5_skin_abund),c("Family","Genus","Species")])
  P5_skin_abund$OTU<-paste(P5_skin_abund$Family,P5_skin_abund$Genus,P5_skin_abund$Species, rownames(P5_skin_abund))
  P5_skin_abund<-P5_skin_abund[,c(1,2,3,4,5,9)]
  P5_skin_top10<-P5_skin_abund[1:10,]
  colnames(P5_skin_top10)<-c("0","2","3","4","5","OTU")
  P5_skin_top10<-melt(P5_skin_top10)
  #Now plot a stacked barplot
  P5_skin_plot<-ggplot(P5_skin_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Patient 5")+ labs(x = "Time point", y="Relative abundance")
  P5_skin_plot
  #make otu table only for P6 wound swabs, and sort by rowsums.  Then convert counts to %, order columns by name, add taxonomy, bind to a single OTU columns with the numerical OTU identifier, subset to top ten and plot over time using ggplot.
  P6sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="P6"]
  P6_skin<-rownames(P6sd)[P6sd[,"sampletype2"]=="diabetic_skin_contra"]
  P6_skin_abund<-data.frame(otu_table(rare_mydata_sub)[,P6_skin])
  P6_skin_abund<-P6_skin_abund[order(-rowSums(P6_skin_abund)),]
  P6_skin_abund<-(P6_skin_abund/30000)*100
  P6_skin_abund<-P6_skin_abund[,order(names(P6_skin_abund))]
  P6_skin_abund<-cbind(P6_skin_abund,tax_table(rare_mydata_sub)[rownames(P6_skin_abund),c("Family","Genus","Species")])
  P6_skin_abund$OTU<-paste(P6_skin_abund$Family,P6_skin_abund$Genus,P6_skin_abund$Species, rownames(P6_skin_abund))
  P6_skin_abund<-P6_skin_abund[,c(1,2,3,4,5,6,10)]
  P6_skin_top10<-P6_skin_abund[1:10,]
  colnames(P6_skin_top10)<-c("0","1","2","3","4","5","OTU")
  P6_skin_top10<-melt(P6_skin_top10)
  #Now plot a stacked barplot
  P6_skin_plot<-ggplot(P6_skin_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Patient 6")+ labs(x = "Time point", y="Relative abundance")
  P6_skin_plot
  #make otu table only for P7 wound swabs, and sort by rowsums.  Then convert counts to %, order columns by name, add taxonomy, bind to a single OTU columns with the numerical OTU identifier, subset to top ten and plot over time using ggplot.
  P7sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="P7"]
  P7_skin<-rownames(P7sd)[P7sd[,"sampletype2"]=="diabetic_skin_contra"]
  P7_skin_abund<-data.frame(otu_table(rare_mydata_sub)[,P7_skin])
  P7_skin_abund<-P7_skin_abund[order(-rowSums(P7_skin_abund)),]
  P7_skin_abund<-(P7_skin_abund/30000)*100
  P7_skin_abund<-P7_skin_abund[,order(names(P7_skin_abund))]
  P7_skin_abund<-cbind(P7_skin_abund,tax_table(rare_mydata_sub)[rownames(P7_skin_abund),c("Family","Genus","Species")])
  P7_skin_abund$OTU<-paste(P7_skin_abund$Family,P7_skin_abund$Genus,P7_skin_abund$Species, rownames(P7_skin_abund))
  P7_skin_abund<-P7_skin_abund[,c(1,2,3,4,5,9)]
  P7_skin_top10<-P7_skin_abund[1:10,]
  colnames(P7_skin_top10)<-c("0","1","2","3","4","OTU")
P7_skin_top10<-melt(P7_skin_top10)
#Now plot a stacked barplot
P7_skin_plot<-ggplot(P7_skin_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Patient 7")+ labs(x = "Time point", y="Relative abundance")
P7_skin_plot
#make otu table only for P8 wound swabs, and sort by rowsums.  Then convert counts to %, order columns by name, add taxonomy, bind to a single OTU columns with the numerical OTU identifier, subset to top ten and plot over time using ggplot.
P8sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="P8"]
P8_skin<-rownames(P8sd)[P8sd[,"sampletype2"]=="diabetic_skin_contra"]
P8_skin_abund<-data.frame(otu_table(rare_mydata_sub)[,P8_skin])
P8_skin_abund<-P8_skin_abund[order(-rowSums(P8_skin_abund)),]
P8_skin_abund<-(P8_skin_abund/30000)*100
P8_skin_abund<-P8_skin_abund[,order(names(P8_skin_abund))]
P8_skin_abund<-cbind(P8_skin_abund,tax_table(rare_mydata_sub)[rownames(P8_skin_abund),c("Family","Genus","Species")])
P8_skin_abund$OTU<-paste(P8_skin_abund$Family,P8_skin_abund$Genus,P8_skin_abund$Species, rownames(P8_skin_abund))
P8_skin_abund<-P8_skin_abund[,c(1,2,3,4,5,9)]
P8_skin_top10<-P8_skin_abund[1:10,]
colnames(P8_skin_top10)<-c("0","1","2","3","5","OTU")
P8_skin_top10<-melt(P8_skin_top10)
#Now plot a stacked barplot
P8_skin_plot<-ggplot(P8_skin_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Patient 8")+ labs(x = "Time point", y="Relative abundance")
P8_skin_plot
#make otu table only for P9 wound swabs, and sort by rowsums.  Then convert counts to %, order columns by name, add taxonomy, bind to a single OTU columns with the numerical OTU identifier, subset to top ten and plot over time using ggplot.
  P9sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="P9"]
  P9_skin<-rownames(P9sd)[P9sd[,"sampletype2"]=="diabetic_skin_contra"]
  P9_skin_abund<-data.frame(otu_table(rare_mydata_sub)[,P9_skin])
  P9_skin_abund<-P9_skin_abund[order(-rowSums(P9_skin_abund)),]
  P9_skin_abund<-(P9_skin_abund/30000)*100
  P9_skin_abund<-P9_skin_abund[,order(names(P9_skin_abund))]
  P9_skin_abund<-cbind(P9_skin_abund,tax_table(rare_mydata_sub)[rownames(P9_skin_abund),c("Family","Genus","Species")])
  P9_skin_abund$OTU<-paste(P9_skin_abund$Family,P9_skin_abund$Genus,P9_skin_abund$Species, rownames(P9_skin_abund))
  P9_skin_abund<-P9_skin_abund[,c(1,2,3,4,5,6,10)]
  P9_skin_top10<-P9_skin_abund[1:10,]
  colnames(P9_skin_top10)<-c("0","1","2","3","4","5","OTU")
  P9_skin_top10<-melt(P9_skin_top10)
  #Now plot a stacked barplot
  P9_skin_plot<-ggplot(P9_skin_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Patient 9")+ labs(x = "Time point", y="Relative abundance")
  P9_skin_plot
  #make otu table only for P10 wound swabs, and sort by rowsums.  Then convert counts to %, order columns by name, add taxonomy, bind to a single OTU columns with the numerical OTU identifier, subset to top ten and plot over time using ggplot.
  P10sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="P10"]
  P10_skin<-rownames(P10sd)[P10sd[,"sampletype2"]=="diabetic_skin_contra"]
  P10_skin_abund<-data.frame(otu_table(rare_mydata_sub)[,P10_skin])
  P10_skin_abund<-P10_skin_abund[order(-rowSums(P10_skin_abund)),]
  P10_skin_abund<-(P10_skin_abund/30000)*100
  P10_skin_abund<-P10_skin_abund[,order(names(P10_skin_abund))]
  P10_skin_abund<-cbind(P10_skin_abund,tax_table(rare_mydata_sub)[rownames(P10_skin_abund),c("Family","Genus","Species")])
  P10_skin_abund$OTU<-paste(P10_skin_abund$Family,P10_skin_abund$Genus,P10_skin_abund$Species, rownames(P10_skin_abund))
  P10_skin_abund<-P10_skin_abund[,c(1,2,3,4,5,9)]
  P10_skin_top10<-P10_skin_abund[1:10,]
  colnames(P10_skin_top10)<-c("0","1","2","3","4","OTU")
  P10_skin_top10<-melt(P10_skin_top10)
  #Now plot a stacked barplot
  P10_skin_plot<-ggplot(P10_skin_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Patient 10")+ labs(x = "Time point", y="Relative abundance")
  P10_skin_plot
  #Now try for CP1
  CP1sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="CP1"]
  CP1_skin<-rownames(CP1sd)
  CP1_skin_abund<-data.frame(otu_table(rare_mydata_sub)[,CP1_skin])
  CP1_skin_abund<-CP1_skin_abund[order(-rowSums(CP1_skin_abund)),]
  CP1_skin_abund<-(CP1_skin_abund/30000)*100
  CP1_skin_abund<-CP1_skin_abund[,order(names(CP1_skin_abund))]
  CP1_skin_abund<-cbind(CP1_skin_abund,tax_table(rare_mydata_sub)[rownames(CP1_skin_abund),c("Family","Genus","Species")])
  CP1_skin_abund$OTU<-paste(CP1_skin_abund$Family,CP1_skin_abund$Genus,CP1_skin_abund$Species, rownames(CP1_skin_abund))
  CP1_skin_abund<-CP1_skin_abund[,c(1,2,3,4,5,6,7,8,9,10,11,15)]
  CP1_skin_top10<-CP1_skin_abund[1:10,]
  colnames(CP1_skin_top10)<-c("0L","0R","1L","1R","2L","2R","3L","4L","4R","5L","5R","OTU")
  CP1_skin_top10<-melt(CP1_skin_top10)
  #Now plot a stacked barplot
  CP1_skin_plot<-ggplot(CP1_skin_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Control 1")+ labs(x = "Time point", y="Relative abundance")
  CP1_skin_plot
  #CP2
  CP2sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="CP2"]
  CP2_skin<-rownames(CP2sd)
  CP2_skin_abund<-data.frame(otu_table(rare_mydata_sub)[,CP2_skin])
  CP2_skin_abund<-CP2_skin_abund[order(-rowSums(CP2_skin_abund)),]
  CP2_skin_abund<-(CP2_skin_abund/30000)*100
  CP2_skin_abund<-CP2_skin_abund[,order(names(CP2_skin_abund))]
  CP2_skin_abund<-cbind(CP2_skin_abund,tax_table(rare_mydata_sub)[rownames(CP2_skin_abund),c("Family","Genus","Species")])
  CP2_skin_abund$OTU<-paste(CP2_skin_abund$Family,CP2_skin_abund$Genus,CP2_skin_abund$Species, rownames(CP2_skin_abund))
  CP2_skin_abund<-CP2_skin_abund[,c(1,2,3,4,5,6,7,8,9,10,14)]
  CP2_skin_top10<-CP2_skin_abund[1:10,]
  colnames(CP2_skin_top10)<-c("0L","0R","1L","1R","2L","2R","4L","4R","5L","5R","OTU")
  CP2_skin_top10<-melt(CP2_skin_top10)
  #Now plot a stacked barplot
  CP2_skin_plot<-ggplot(CP2_skin_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Control 2")+ labs(x = "Time point", y="Relative abundance")
  CP2_skin_plot
  #CP3
  CP3sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="CP3"]
  CP3_skin<-rownames(CP3sd)
  CP3_skin_abund<-data.frame(otu_table(rare_mydata_sub)[,CP3_skin])
  CP3_skin_abund<-CP3_skin_abund[order(-rowSums(CP3_skin_abund)),]
  CP3_skin_abund<-(CP3_skin_abund/30000)*100
  CP3_skin_abund<-CP3_skin_abund[,order(names(CP3_skin_abund))]
  CP3_skin_abund<-cbind(CP3_skin_abund,tax_table(rare_mydata_sub)[rownames(CP3_skin_abund),c("Family","Genus","Species")])
  CP3_skin_abund$OTU<-paste(CP3_skin_abund$Family,CP3_skin_abund$Genus,CP3_skin_abund$Species, rownames(CP3_skin_abund))
  CP3_skin_abund<-CP3_skin_abund[,c(1,2,3,4,5,6,7,8,12)]
  CP3_skin_top10<-CP3_skin_abund[1:10,]
  colnames(CP3_skin_top10)<-c("1L","1R","2L","2R","3L","3R","4L","4R","OTU")
  CP3_skin_top10<-melt(CP3_skin_top10)
  #Now plot a stacked barplot
  CP3_skin_plot<-ggplot(CP3_skin_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Control 3")+ labs(x = "Time point", y="Relative abundance")
  CP3_skin_plot
  #CP4
  CP4sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="CP4"]
  CP4_skin<-rownames(CP4sd)
  CP4_skin_abund<-data.frame(otu_table(rare_mydata_sub)[,CP4_skin])
  CP4_skin_abund<-CP4_skin_abund[order(-rowSums(CP4_skin_abund)),]
  CP4_skin_abund<-(CP4_skin_abund/30000)*100
  CP4_skin_abund<-CP4_skin_abund[,order(names(CP4_skin_abund))]
  CP4_skin_abund<-cbind(CP4_skin_abund,tax_table(rare_mydata_sub)[rownames(CP4_skin_abund),c("Family","Genus","Species")])
  CP4_skin_abund$OTU<-paste(CP4_skin_abund$Family,CP4_skin_abund$Genus,CP4_skin_abund$Species, rownames(CP4_skin_abund))
  CP4_skin_abund<-CP4_skin_abund[,c(1,2,3,4,5,6,7,8,9,10,14)]
  CP4_skin_top10<-CP4_skin_abund[1:10,]
  colnames(CP4_skin_top10)<-c("0L","0R","2L","2R","3L","3R","4L","4R","5L","5R","OTU")
  CP4_skin_top10<-melt(CP4_skin_top10)
  #Now plot a stacked barplot
  CP4_skin_plot<-ggplot(CP4_skin_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Control 4")+ labs(x = "Time point", y="Relative abundance")
  CP4_skin_plot
  #CP5
  CP5sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="CP5"]
  CP5_skin<-rownames(CP5sd)
  CP5_skin_abund<-data.frame(otu_table(rare_mydata_sub)[,CP5_skin])
  CP5_skin_abund<-CP5_skin_abund[order(-rowSums(CP5_skin_abund)),]
  CP5_skin_abund<-(CP5_skin_abund/30000)*100
  CP5_skin_abund<-CP5_skin_abund[,order(names(CP5_skin_abund))]
  CP5_skin_abund<-cbind(CP5_skin_abund,tax_table(rare_mydata_sub)[rownames(CP5_skin_abund),c("Family","Genus","Species")])
  CP5_skin_abund$OTU<-paste(CP5_skin_abund$Family,CP5_skin_abund$Genus,CP5_skin_abund$Species, rownames(CP5_skin_abund))
  CP5_skin_abund<-CP5_skin_abund[,c(1,2,3,4,5,6,7,8,9,10,11,12,16)]
  CP5_skin_top10<-CP5_skin_abund[1:10,]
  colnames(CP5_skin_top10)<-c("0L","0R","1L","1R","2L","2R","3L","3R","4L","4R","5L","5R","OTU")
  CP5_skin_top10<-melt(CP5_skin_top10)
  #Now plot a stacked barplot
  CP5_skin_plot<-ggplot(CP5_skin_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Control 5")+ labs(x = "Time point", y="Relative abundance")
  CP5_skin_plot
  #CP6
  CP6sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="CP6"]
  CP6_skin<-rownames(CP6sd)
  CP6_skin_abund<-data.frame(otu_table(rare_mydata_sub)[,CP6_skin])
  CP6_skin_abund<-CP6_skin_abund[order(-rowSums(CP6_skin_abund)),]
  CP6_skin_abund<-(CP6_skin_abund/30000)*100
  CP6_skin_abund<-CP6_skin_abund[,order(names(CP6_skin_abund))]
  CP6_skin_abund<-cbind(CP6_skin_abund,tax_table(rare_mydata_sub)[rownames(CP6_skin_abund),c("Family","Genus","Species")])
  CP6_skin_abund$OTU<-paste(CP6_skin_abund$Family,CP6_skin_abund$Genus,CP6_skin_abund$Species, rownames(CP6_skin_abund))
  CP6_skin_abund<-CP6_skin_abund[,c(1,2,3,4,5,6,7,8,9,13)]
  CP6_skin_top10<-CP6_skin_abund[1:10,]
  colnames(CP6_skin_top10)<-c("1L","1R","2L","2R","3L","3R","4L","4R","5L","OTU")
  CP6_skin_top10<-melt(CP6_skin_top10)
  #Now plot a stacked barplot
  CP6_skin_plot<-ggplot(CP6_skin_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Control 6")+ labs(x = "Time point", y="Relative abundance")
  CP6_skin_plot
  #CP7
  CP7sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="CP7"]
  CP7_skin<-rownames(CP7sd)
  CP7_skin_abund<-data.frame(otu_table(rare_mydata_sub)[,CP7_skin])
  CP7_skin_abund<-CP7_skin_abund[order(-rowSums(CP7_skin_abund)),]
  CP7_skin_abund<-(CP7_skin_abund/30000)*100
  CP7_skin_abund<-CP7_skin_abund[,order(names(CP7_skin_abund))]
  CP7_skin_abund<-cbind(CP7_skin_abund,tax_table(rare_mydata_sub)[rownames(CP7_skin_abund),c("Family","Genus","Species")])
  CP7_skin_abund$OTU<-paste(CP7_skin_abund$Family,CP7_skin_abund$Genus,CP7_skin_abund$Species, rownames(CP7_skin_abund))
  CP7_skin_abund<-CP7_skin_abund[,c(1,2,3,4,5,6,7,8,9,13)]
  CP7_skin_top10<-CP7_skin_abund[1:10,]
  colnames(CP7_skin_top10)<-c("0R","0L","1L","2L","3L","3R","4L","5L","5R","OTU")
  CP7_skin_top10<-melt(CP7_skin_top10)
  #Now plot a stacked barplot
  CP7_skin_plot<-ggplot(CP7_skin_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Control 7")+ labs(x = "Time point", y="Relative abundance")
  CP7_skin_plot
  #CP9
  CP9sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="CP9"]
  CP9_skin<-rownames(CP9sd)
  CP9_skin_abund<-data.frame(otu_table(rare_mydata_sub)[,CP9_skin])
  CP9_skin_abund<-CP9_skin_abund[order(-rowSums(CP9_skin_abund)),]
  CP9_skin_abund<-(CP9_skin_abund/30000)*100
  CP9_skin_abund<-CP9_skin_abund[,order(names(CP9_skin_abund))]
  CP9_skin_abund<-cbind(CP9_skin_abund,tax_table(rare_mydata_sub)[rownames(CP9_skin_abund),c("Family","Genus","Species")])
  CP9_skin_abund$OTU<-paste(CP9_skin_abund$Family,CP9_skin_abund$Genus,CP9_skin_abund$Species, rownames(CP9_skin_abund))
  CP9_skin_abund<-CP9_skin_abund[,c(1,2,3,4,5,6,7,8,9,10,11,12,16)]
  CP9_skin_top10<-CP9_skin_abund[1:10,]
  colnames(CP9_skin_top10)<-c("0L","0R","1L","1R","2L","2R","3L","3R","4L","4R","5L","5R","OTU")
  CP9_skin_top10<-melt(CP9_skin_top10)
  #Now plot a stacked barplot
  CP9_skin_plot<-ggplot(CP9_skin_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Control 9")+ labs(x = "Time point", y="Relative abundance")
  CP9_skin_plot

```
Wound taxa plots top 10 and 20 overall
```{r, echo=FALSE}
#Create vector of sample names for wounds and wond debridement
diabetic_woundswab_samples<-rownames(sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"sampletype2"]=="wound_swab"])
diabetic_wounddeb_samples<-rownames(sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"sampletype2"]=="wound_deb"])
#Now apply the mean function only to columns of the OTU table that correspond to a sample from each of the vectors of groups sample names. i.e. get mean abundance for each OTU for each of the sample groups defined above.
mean_abund_woundswab_otus<-as.matrix(apply(otu_table(rare_mydata_sub)[,diabetic_woundswab_samples],1, FUN=mean))
mean_abund_wounddeb_otus<-as.matrix(apply(otu_table(rare_mydata_sub)[,diabetic_wounddeb_samples],1, FUN=mean))
#Now bind those results into a new table and define row (OTU) and column (sample) names.
average_abund_wounds<-as.matrix(cbind(mean_abund_woundswab_otus,mean_abund_wounddeb_otus))
rownames(average_abund_wounds)<-rownames(otu_table(rare_mydata_sub))
colnames(average_abund_wounds)<-c("wound swabs","wound debridment")
#Now sort table by the 20 most abundant across all sample types, after converting counts to propotions.
average_abund_wounds<-((average_abund_wounds/30000)*100)
average_abund_wounds<-average_abund_wounds[order(-rowSums(average_abund_wounds)),]
top20wounds<-average_abund_wounds[1:20,]
top20wounds<-data.frame(top20wounds)
top20wounds<-cbind(top20wounds,tax_table(rare_mydata_sub)[rownames(top20wounds),])
top20wounds$OTU<-paste(top20wounds$Family, top20wounds$Genus,rownames(top20wounds))
top20wounds_melt<-top20wounds[,c(1,2,10)]
#Now melt data to the right format for ggplot
top20wounds_melt<-melt(top20wounds_melt)
#Now plot a stacked barplot
ggplot(top20wounds_melt, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_hue(l=50)+ggtitle("Top 20 OTUs in Wounds")+labs(x = "", y="Relative abundance (%)")+ guides(fill=guide_legend(keywidth=0.5, keyheight=0.5))
#Now the top 10
top10wounds<-average_abund_wounds[1:10,]
top10wounds<-data.frame(top10wounds)
top10wounds<-cbind(top10wounds,tax_table(rare_mydata_sub)[rownames(top10wounds),])
top10wounds$OTU<-paste(top10wounds$Family, top10wounds$Genus,rownames(top10wounds))
top10wounds_melt<-top10wounds[,c(1,2,10)]
#Now melt data to the right format for ggplot
top10wounds_melt<-melt(top10wounds_melt)
#Now plot a stacked barplot
ggplot(top10wounds_melt, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+ggtitle("Top 10 OTUs in Wounds")+labs(x = "", y="Relative abundance (%)")+ theme(legend.text=element_text(size=8), legend.title=element_text(size=10),legend.key.size = unit(0.5, "cm"))+scale_fill_brewer(palette="Paired")

```

Wound swabs per patient over time
```{r, echo=FALSE}
#create vectors of sample names for wound swabs for each patient
P1sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="P1"]
P1_wounds<-rownames(P1sd)[P1sd[,"sampletype2"]=="wound_swab"]
P2sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="P2"]
P2_wounds<-rownames(P2sd)[P2sd[,"sampletype2"]=="wound_swab"]
P5sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="P5"]
P5_wounds<-rownames(P5sd)[P5sd[,"sampletype2"]=="wound_swab"]
P6sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="P6"]
P6_wounds<-rownames(P6sd)[P6sd[,"sampletype2"]=="wound_swab"]
P7sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="P7"]
P7_wounds<-rownames(P7sd)[P7sd[,"sampletype2"]=="wound_swab"]
P8sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="P8"]
P8_wounds<-rownames(P8sd)[P8sd[,"sampletype2"]=="wound_swab"]
P9sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="P9"]
P9_wounds<-rownames(P9sd)[P9sd[,"sampletype2"]=="wound_swab"]
P10sd<-sample_data(rare_mydata_sub)[sample_data(rare_mydata_sub)[,"subject"]=="P10"]
P10_wounds<-rownames(P10sd)[P10sd[,"sampletype2"]=="wound_swab"]
#make otu table only for P1 wound swabs, and sort by rowsums.  Then convert counts to %, order columns by name, add taxonomy, bind to a single OTU columns with the numerical OTU identifier, subset to top ten and plot over time using ggplot.
P1_wounds_abund<-data.frame(otu_table(rare_mydata_sub)[,P1_wounds])
P1_wounds_abund<-P1_wounds_abund[order(-rowSums(P1_wounds_abund)),]
P1_wounds_abund<-(P1_wounds_abund/30000)*100
P1_wounds_abund<-P1_wounds_abund[,order(names(P1_wounds_abund))]
P1_wounds_abund<-cbind(P1_wounds_abund,tax_table(rare_mydata_sub)[rownames(P1_wounds_abund),c("Family","Genus","Species")])
P1_wounds_abund$OTU<-paste(P1_wounds_abund$Family,P1_wounds_abund$Genus,P1_wounds_abund$Species, rownames(P1_wounds_abund))
P1_wounds_abund<-P1_wounds_abund[,c(1,2,3,4,5,6,10)]
P1_wounds_top10<-P1_wounds_abund[1:10,]
colnames(P1_wounds_top10)<-c("0","1","2","3","4","5","OTU")
P1_wounds_top10<-melt(P1_wounds_top10)
#create data.frame of wound area data with compatible variables and merge with top 10 data. Wound size represented as a % of largest area measured to be compatible with axis for taxa plotting
P1_size<-data.frame(variable=c("0","1","2","3","4","5"), area=c(100,58.2,21.8,21.8,12.7,16.4))
P1_wounds_top10<-merge(P1_wounds_top10,P1_size)
#Now plot a stacked barplot
P1_wound_plot<-ggplot(P1_wounds_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Patient 1")+ labs(x = "Time point", y="Relative abundance and wound area(%)")
P1_wound_plot
#add wound size to plot
P1_plot<-P1_wound_plot+geom_line(aes(y=area), colour = "red", group=1)
P1_plot
#Still need to figure our best colour scale and labels - will do this later.
#Repeat for patient 2
P2_wounds_abund<-data.frame(otu_table(rare_mydata_sub)[,P2_wounds])
P2_wounds_abund<-P2_wounds_abund[order(-rowSums(P2_wounds_abund)),]
P2_wounds_abund<-(P2_wounds_abund/30000)*100
P2_wounds_abund<-P2_wounds_abund[,order(names(P2_wounds_abund))]
P2_wounds_abund<-cbind(P2_wounds_abund,tax_table(rare_mydata_sub)[rownames(P2_wounds_abund),c("Family","Genus","Species")])
P2_wounds_abund$OTU<-paste(P2_wounds_abund$Family,P2_wounds_abund$Genus,P2_wounds_abund$Species, rownames(P2_wounds_abund))
P2_wounds_abund<-P2_wounds_abund[,c(1,2,3,7)]
P2_wounds_top10<-P2_wounds_abund[1:10,]
colnames(P2_wounds_top10)<-c("0","3","5","OTU")
P2_wounds_top10<-melt(P2_wounds_top10)
P2_size<-data.frame(variable=c("0","3","5"), area=c(66.7,33.3,100))
P2_wounds_top10<-merge(P2_wounds_top10,P2_size)
#Now plot a stacked barplot
P2_wound_plot<-ggplot(P2_wounds_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Patient 2")+ labs(x = "Time point", y="Relative abundance and wound area(%)")
P2_wound_plot
#add wound size to plot
P2_plot<-P2_wound_plot+geom_line(aes(y=area), colour = "red", group=1)
P2_plot
#Now repeat for P5
P5_wounds_abund<-data.frame(otu_table(rare_mydata_sub)[,P5_wounds])
P5_wounds_abund<-P5_wounds_abund[order(-rowSums(P5_wounds_abund)),]
P5_wounds_abund<-(P5_wounds_abund/30000)*100
P5_wounds_abund<-P5_wounds_abund[,order(names(P5_wounds_abund))]
P5_wounds_abund<-cbind(P5_wounds_abund,tax_table(rare_mydata_sub)[rownames(P5_wounds_abund),c("Family","Genus","Species")])
P5_wounds_abund$OTU<-paste(P5_wounds_abund$Family,P5_wounds_abund$Genus,P5_wounds_abund$Species, rownames(P5_wounds_abund))
P5_wounds_abund<-P5_wounds_abund[,c(1,2,3,4,5,6,10)]
P5_wounds_top10<-P5_wounds_abund[1:10,]
colnames(P5_wounds_top10)<-c("0","1","2","3","4","5","OTU")
P5_wounds_top10<-melt(P5_wounds_top10)
P5_size<-data.frame(variable=c("0","1","2","3","4","5"), area=c(12.5,12.5,100,8.3,31.3,52.1))
P5_wounds_top10<-merge(P5_wounds_top10,P5_size)
#Now plot a stacked barplot
P5_wound_plot<-ggplot(P5_wounds_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Patient 5")+ labs(x = "Time point", y="Relative abundance and wound area(%)")
P5_wound_plot
#add wound size to plot
P5_plot<-P5_wound_plot+geom_line(aes(y=area), colour = "red", group=1)
P5_plot
#Repeat for P6
P6_wounds_abund<-data.frame(otu_table(rare_mydata_sub)[,P6_wounds])
P6_wounds_abund<-P6_wounds_abund[order(-rowSums(P6_wounds_abund)),]
P6_wounds_abund<-(P6_wounds_abund/30000)*100
P6_wounds_abund<-P6_wounds_abund[,order(names(P6_wounds_abund))]
P6_wounds_abund<-cbind(P6_wounds_abund,tax_table(rare_mydata_sub)[rownames(P6_wounds_abund),c("Family","Genus","Species")])
P6_wounds_abund$OTU<-paste(P6_wounds_abund$Family,P6_wounds_abund$Genus,P6_wounds_abund$Species, rownames(P6_wounds_abund))
P6_wounds_abund<-P6_wounds_abund[,c(1,2,3,4,5,9)]
P6_wounds_top10<-P6_wounds_abund[1:10,]
colnames(P6_wounds_top10)<-c("0","1","2","3","4","OTU")
P6_wounds_top10<-melt(P6_wounds_top10)
P6_size<-data.frame(variable=c("0","1","2","3","4"), area=c(80,100,80,60,0))
P6_wounds_top10<-merge(P6_wounds_top10,P6_size)
#Now plot a stacked barplot
P6_wound_plot<-ggplot(P6_wounds_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Patient 6")+ labs(x = "Time point", y="Relative abundance and wound area(%)")
P6_wound_plot
#add wound size to plot
P6_plot<-P6_wound_plot+geom_line(aes(y=area), colour = "red", group=1)
P6_plot
#Repeat for P7
P7_wounds_abund<-data.frame(otu_table(rare_mydata_sub)[,P7_wounds])
P7_wounds_abund<-P7_wounds_abund[order(-rowSums(P7_wounds_abund)),]
P7_wounds_abund<-(P7_wounds_abund/30000)*100
P7_wounds_abund<-P7_wounds_abund[,order(names(P7_wounds_abund))]
P7_wounds_abund<-cbind(P7_wounds_abund,tax_table(rare_mydata_sub)[rownames(P7_wounds_abund),c("Family","Genus","Species")])
P7_wounds_abund$OTU<-paste(P7_wounds_abund$Family,P7_wounds_abund$Genus,P7_wounds_abund$Species, rownames(P7_wounds_abund))
P7_wounds_abund<-P7_wounds_abund[,c(1,2,3,4,8)]
P7_wounds_top10<-P7_wounds_abund[1:10,]
colnames(P7_wounds_top10)<-c("0","1","2","3","OTU")
P7_wounds_top10<-melt(P7_wounds_top10)
P7_size<-data.frame(variable=c("0","1","2","3"), area=c(100,91.7,25,0))
P7_wounds_top10<-merge(P7_wounds_top10,P7_size)
#Now plot a stacked barplot
P7_wound_plot<-ggplot(P7_wounds_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ggtitle("Patient 7")+ labs(x = "Time point", y="Relative abundance and wound area(%)")
P7_wound_plot
#add wound size to plot
P7_plot<-P7_wound_plot+geom_line(aes(y=area), colour = "red", group=1)
P7_plot
#Repeat for P8
P8_wounds_abund<-data.frame(otu_table(rare_mydata_sub)[,P8_wounds])
P8_wounds_abund<-P8_wounds_abund[order(-rowSums(P8_wounds_abund)),]
P8_wounds_abund<-(P8_wounds_abund/30000)*100
P8_wounds_abund<-P8_wounds_abund[,order(names(P8_wounds_abund))]
P8_wounds_abund<-cbind(P8_wounds_abund,tax_table(rare_mydata_sub)[rownames(P8_wounds_abund),c("Family","Genus","Species")])
P8_wounds_abund$OTU<-paste(P8_wounds_abund$Family,P8_wounds_abund$Genus,P8_wounds_abund$Species, rownames(P8_wounds_abund))
P8_wounds_abund<-P8_wounds_abund[,c(1,2,3,4,5,9)]
P8_wounds_top10<-P8_wounds_abund[1:10,]
colnames(P8_wounds_top10)<-c("0","1","2","3","5","OTU")
P8_wounds_top10<-melt(P8_wounds_top10)
P8_size<-data.frame(variable=c("0","1","2","3","5"), area=c(33.3,100,72.9,83.3,83.3))
P8_wounds_top10<-merge(P8_wounds_top10,P8_size)
#Now plot a stacked barplot
P8_wound_plot<-ggplot(P8_wounds_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ labs(x = "Time point", y="Relative abundance and wound area(%)")+ggtitle("Patient 8")
P8_wound_plot
#add wound size to plot
P8_plot<-P8_wound_plot+geom_line(aes(y=area), colour = "red", group=1)
P8_plot
#Repeat for P9
P9_wounds_abund<-data.frame(otu_table(rare_mydata_sub)[,P9_wounds])
P9_wounds_abund<-P9_wounds_abund[order(-rowSums(P9_wounds_abund)),]
P9_wounds_abund<-(P9_wounds_abund/30000)*100
P9_wounds_abund<-P9_wounds_abund[,order(names(P9_wounds_abund))]
P9_wounds_abund<-cbind(P9_wounds_abund,tax_table(rare_mydata_sub)[rownames(P9_wounds_abund),c("Family","Genus","Species")])
P9_wounds_abund$OTU<-paste(P9_wounds_abund$Family,P9_wounds_abund$Genus,P9_wounds_abund$Species, rownames(P9_wounds_abund))
P9_wounds_abund<-P9_wounds_abund[,c(1,2,3,4,5,9)]
P9_wounds_top10<-P9_wounds_abund[1:10,]
colnames(P9_wounds_top10)<-c("0","1","2","4","5","OTU")
P9_wounds_top10<-melt(P9_wounds_top10)
P9_size<-data.frame(variable=c("0","1","2","3","4","5"), area=c(90.9,95,100,100,100,77.7))
P9_wounds_top10<-merge(P9_wounds_top10,P9_size)
#Now plot a stacked barplot
P9_wound_plot<-ggplot(P9_wounds_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ labs(x = "Time point", y="Relative abundance and wound area(%)")+ggtitle("Patient 9")
P9_wound_plot
#add wound size to plot
P9_plot<-P9_wound_plot+geom_line(aes(y=area), colour = "red", group=1)
P9_plot
#Repeat for P10
P10_wounds_abund<-data.frame(otu_table(rare_mydata_sub)[,P10_wounds])
P10_wounds_abund<-P10_wounds_abund[order(-rowSums(P10_wounds_abund)),]
P10_wounds_abund<-(P10_wounds_abund/30000)*100
P10_wounds_abund<-P10_wounds_abund[,order(names(P10_wounds_abund))]
P10_wounds_abund<-cbind(P10_wounds_abund,tax_table(rare_mydata_sub)[rownames(P10_wounds_abund),c("Family","Genus","Species")])
P10_wounds_abund$OTU<-paste(P10_wounds_abund$Family,P10_wounds_abund$Genus,P10_wounds_abund$Species, rownames(P10_wounds_abund))
P10_wounds_abund<-P10_wounds_abund[,c(1,2,3,4,5,9)]
P10_wounds_top10<-P10_wounds_abund[1:10,]
colnames(P10_wounds_top10)<-c("0","1","2","3","4","OTU")
P10_wounds_top10<-melt(P10_wounds_top10)
P10_size<-data.frame(variable=c("0","1","2","3","4"), area=c(97.2,88.9,91.7,100,44.4))
P10_wounds_top10<-merge(P10_wounds_top10,P10_size)
#Now plot a stacked barplot
P10_wound_plot<-ggplot(P10_wounds_top10, aes(x=variable, y=value, fill=OTU))+geom_bar(stat="identity")+scale_fill_brewer(palette="Paired")+ labs(x = "Time point", y="Relative abundance and wound area(%)")+ggtitle("Patient 10")
P10_wound_plot
#add wound size to plot
P10_plot<-P10_wound_plot+geom_line(aes(y=area), colour = "red", group=1)
P10_plot
```
Weighted unifrac distance with PCoA ordination and plotting on rarefied otu table
```{r, echo=FALSE}
diaball_dist<-phyloseq::distance(rare_mydata_sub, method="unifrac", type="samples", weighted=TRUE)
diaball_ord<-ordinate(rare_mydata_sub, method="PCoA", distance="diaball_dist", type="samples")
```
```{r, echo=FALSE, fig.width=10, fig.height=4}
plot_ordination(rare_mydata_sub, diaball_ord, type="samples", color="sampletype1", title="PCoA of weighted unifrac distances between wounds, diabetic skin,\nand non-diabetic skin\nrarefied OTU table")
#diaball_dist_jc<-phyloseq::distance(rare_mydata_sub, method="jaccard", type="samples")
#diaball_ord_jc<-ordinate(rare_mydata_sub, method="PCoA", distance="diaball_dist_jc", type="samples")
#plot_ordination(rare_mydata_sub, diaball_ord_jc, type="samples", color="sampletype1", title="PCoA of Jaccard distances between wounds, diabetic skin,\nand non-diabetic skin\nrarefied OTU table")
```
Subset to skin only and repeat
```{r, echo=FALSE}
rare_skin<-subset_samples(rare_mydata_sub, sampletype1%in%c("control_skin","diabetic_skin"))
rare_skin_dist<-phyloseq::distance(rare_skin, method="unifrac", type="samples", weighted=TRUE)
rare_skin_ord<-ordinate(rare_skin, method="PCoA", distance="rare_skin_dist", type="samples")
```
```{r, echo=FALSE, fig.width=10, fig.height=4}
plot_ordination(rare_skin, rare_skin_ord, type="samples", color="sampletype1", title="PCoA of weighted unifrac distances between diabetic skin,\nand non-diabetic skin\n rarefied OTU table")
#rare_skin_dist_jc<-phyloseq::distance(rare_skin, method="jaccard", type="samples")
#rare_skin_ord_jc<-ordinate(rare_skin, method="PCoA", distance="rare_skin_dist_jc", type="samples")
#plot_ordination(rare_skin, rare_skin_ord_jc, type="samples", color="sampletype1", title="PCoA of Jaccard distances between diabetic skin,\nand non-diabetic skin\n rarefied OTU table")
#```
#Try Horn-Morista
#```{r}
#diaball_dist_hm <-phyloseq::distance(rare_mydata_sub, method="horn", type="samples")
#diaball_ord_hm <- ordinate(rare_mydata_sub, method="PCoA", distance="diaball_dist_hm", type="samples")
#```
#```{r fig.width=10, fig.height=4}
#plot_ordination(rare_mydata_sub, diaball_ord_hm, type="samples", color="sampletype1", title="PCoA of Horn-Morista distances between wounds, diabetic skin,\nand non-diabetic skin\n rarefied OTU table")
#```
#```{r}
#rare_skin_dist_hm<-phyloseq::distance(rare_skin, method="horn", type="samples")
#rare_skin_ord_hm<-ordinate(rare_skin, method="PCoA", distance="rare_skin_dis#t_hm", type="samples")
#```
#```{r fig.width=10, fig.height=4}
#plot_ordination(rare_skin, rare_skin_ord_hm, type="samples", color="sampletype1", title="PCoA of Horn-Morista distances between diabetic skin,\nand non-diabetic skin\n rarefied OTU table")
#plot_ordination(rare_skin, rare_skin_ord_hm, type="samples", color="subject", shape="sampletype1", title="PCoA of Horn-Morista distances between diabetic skin,\nand non-diabetic skin\n rarefied OTU table")
```
Subset to wounds only and repeat
```{r, echo=FALSE}
rare_wounds<-subset_samples(rare_mydata_sub, sampletype1=="wound")
rare_wounds_dist<-phyloseq::distance(rare_wounds, method="unifrac", type="samples", weighted=TRUE)
rare_wounds_ord<-ordinate(rare_wounds, method="PCoA", distance="rare_wounds_dist", type="samples")
```
```{r, echo=FALSE, fig.width=10, fig.height=4}
plot_ordination(rare_wounds, rare_wounds_ord, type="samples", color="subject", title="PCoA of weighted unifrac distances between diabetic wounds\n rarefied OTU table")+ geom_path()+scale_fill_brewer(palette="Paired")
```
Repeat with diabetic skin and wounds
```{r}
rare_diabetes<-subset_samples(rare_mydata_sub, sampletype1%in%c("wound","diabetic_skin"))
rare_diabetes_dist<-phyloseq::distance(rare_diabetes, method="unifrac", type="samples", weighted=TRUE)
rare_diabetes_ord<-ordinate(rare_diabetes, method="PCoA", distance="rare_diabetes_dist", type="samples")
```
```{r, echo=FALSE, fig.width=10, fig.height=4}
plot_ordination(rare_diabetes, rare_diabetes_ord, type="samples", color="subject", shape="sampletype1", title="PCoA of weighted unifrac distances between diabetic wounds\n and skin, rarefied OTU table")
```
Permanova tests for signficance on rarefied otu distance analysis
```{r, echo=FALSE}
#PERMANOVA-create a file of factors for only the samples in the analysis
mapping<- sample_data(rare_skin)
health<- mapping$health
subject<- mapping$subject
#PERMANOVA analysis
library("vegan")
perm_rare_unifrac<-adonis(rare_skin_dist ~ health/subject, data=as(sample_data(rare_skin),"data.frame"))
perm_rare_unifrac
#perm_rare_jaccard<-adonis(rare_skin_dist_jc ~ health/subject, data=as(sample_data(rare_skin),"data.frame"))
#perm_rare_jaccard
#perm_rare_hm<-adonis(rare_skin_dist_hm ~ health/subject, data=as(sample_data(rare_skin),"data.frame"))
#perm_rare_hm
```
Variance Stabilizing Transformation, Weighted Unifrac and Ordination
```{r, echo=FALSE}
library("DESeq2")
diaball_dds <- phyloseq_to_deseq2(mydata_sub, ~1)
#Filter OTUs present in only one sample 
#What about otus present in less than 10% of samples?
diaball_filtered_dds <- diaball_dds[ rowSums(counts(diaball_dds)) > 1, ]
#estaimte size factors and dispersion in DESeq2 
diaball_filtered_dds <- estimateSizeFactors(diaball_filtered_dds)
diaball_filtered_dds <- estimateDispersions(diaball_filtered_dds, fitType = "local")
#Create new phyloseq object from mydata_sub and populate the otu table with the vst values
diaball_vst <-mydata_sub
otu_table(diaball_vst)<- otu_table(getVarianceStabilizedData(diaball_filtered_dds), taxa_are_rows = TRUE)
#Remove -ive values in matrix, replace with 0
matrix_vst_otu <- otu_table(diaball_vst, taxa_are_rows=TRUE)
matrix_vst_otu[matrix_vst_otu<0]=0
otu_table(diaball_vst)<-otu_table(matrix_vst_otu, taxa_are_rows=TRUE)
#weighted unifrac and ordination
diaball_vst_dist<-phyloseq::distance(diaball_vst, method="unifrac", type="samples", weighted=TRUE)
diaball_vst_ord<-ordinate(diaball_vst, method="PCoA", distance="diaball_vst_dist", type="samples")
```
```{r, echo=FALSE, fig.width=10, fig.height=4}
plot_ordination(diaball_vst, diaball_vst_ord, type="samples", color="sampletype1", title="PCoA of weighted unifrac distances between wounds, diabetic skin,\nand non-diabetic skin\nvst transformed OTU table")
```
Subset to skin only and repeat
```{r, echo=FALSE}
#subset to skin only and repeat
skin_vst<-subset_samples(diaball_vst, sampletype1%in%c("control_skin", "diabetic_skin"))
skin_vst_dist<-phyloseq::distance(skin_vst, method="unifrac", type="samples", weighted=TRUE)
skin_vst_ord<-phyloseq::ordinate(skin_vst, method="PCoA", distance="skin_vst_dist")
```
```{r, echo=FASLE, fig.width=10, fig.height=4}
plot_ordination(skin_vst, skin_vst_ord, type="samples", color="sampletype1", title="PCoA of weighted unifrac distances between\ndiabetic skin and non-diabetic skin\nvst transformed OTU table")+theme(legend.title=element_blank())
Figure3<-plot_ordination(skin_vst, skin_vst_ord, type="samples", color="sampletype1", title="PCoA of weighted unifrac distances between\ndiabetic skin and non-diabetic skin")+theme(legend.title=element_blank())
####Trying to put 95% confidence interval elipses on plot using instructions from here https://www.biostars.org/p/126454/, but can't get it to work - works on CCA plots but not Unifrac PCoA plots according to the example..
#Get data from plot
#skin_vst_ord_data<-(plot_ordination(skin_vst, skin_vst_ord, type="samples", color="sampletype1", title="PCoA of weighted unifrac distances between\ndiabetic skin and non-diabetic skin\nvst transformed OTU table"))$data
#skin_vst_ord_data
# add a fake variable to the experiment
#sample_data(skin_vst)$fake <- 1:nsamples(skin_vst)
# use the envfit function of vegan
#ord_ef_skin_vst <- envfit(skin_vst_ord ~ fake,permu=1000,data = as(sample_data(skin_vst),"data.frame"))
#Use Jaccard and Horn-Morista distances
#```{r}
#diaball_vst_dist_jc<-phyloseq::distance(diaball_vst, method="jaccard", type="samples")
#diaball_vst_ord_jc<-ordinate(diaball_vst, method="PCoA", distance="diaball_vst_dist_jc", type="samples")
#```
#```{r fig.width=10, fig.height=4}
#plot_ordination(diaball_vst, diaball_vst_ord_jc, type="samples", color="sampletype1", title="PCoA of Jaccard distances between wounds, diabetic skin,\nand non-diabetic skin\nvst transformed OTU table")
#```
#```{r}
#repeat on skin only subset
#skin_vst_dist_jc<-phyloseq::distance(skin_vst, method="jaccard", type="samples")
#skin_vst_ord_jc<-phyloseq::ordinate(skin_vst, method="PCoA", distance#="skin_vst_dist_jc")
#```
#```{r fig.width=10, fig.height=4}
#plot_ordination(skin_vst, skin_vst_ord_jc, type="samples", color="sampletype1", title="PCoA of Jaccard distances between\ndiabetic skin and non-diabetic skin\nvst transformed OTU table")
#```
#```{r}
#diaball_dist_hm<-phyloseq::distance(diaball_vst, method="horn", type="samples")
#diaball_ord_hm<-ordinate(diaball_vst, method="PCoA", distance="diaball_dist_hm", type="samples")
#```{r fig.width=10, fig.height=4}
#plot_ordination(diaball_vst, diaball_ord_hm, type="samples", color="sampletype1", title="PCoA of Horn-Morista distances between wounds, diabetic skin,\nand non-diabetic skin\nvst transformed OTU table")
#```
#```{r}
#subset to skin only and repeat
#skin_vst_dist_hm<-phyloseq::distance(skin_vst, method="horn", type="samples")
#skin_vst_ord_hm<-phyloseq::ordinate(skin_vst, method="PCoA", distance="skin_vst_dist_hm")
#```
#```{r fig.width=10, fig.height=4}
#plot_ordination(skin_vst, skin_vst_ord_hm, type="samples", color="sampletype1", title="PCoA of Horn-Morista distances between\ndiabetic skin and non-diabetic skin\nvst transformed OTU table")
```
Subset to wounds only and repeat
```{r, echo=FALSE}
wounds_vst<-subset_samples(diaball_vst, sampletype1=="wound")
wounds_vst_dist<-phyloseq::distance(wounds_vst, method="unifrac", type="samples", weighted=TRUE)
wounds_vst_ord<-phyloseq::ordinate(wounds_vst, method="PCoA", distance="wounds_vst_dist")
```
```{r, echo=FASLE, fig.width=10, fig.height=4}
plot_ordination(wounds_vst, wounds_vst_ord, type="samples", color="subject",shape="sampletype2", title="PCoA of weighted unifrac distances between diabetic wounds\nvst transformed OTU table")+theme(legend.title=element_blank())+scale_fill_brewer(palette="Paired")

```
Permanova statistics on vst otu table distance calculations
```{r, echo=FALSE}
perm_vst_unifrac<-adonis(skin_vst_dist ~ health/subject, data=as(sample_data(skin_vst),"data.frame"))
perm_vst_unifrac
#perm_vst_jc<-adonis(skin_vst_dist_jc ~ health/subject, data=as(sample_data(skin_vst),"data.frame"))
#perm_vst_jc
#perm_vst_hm<-adonis(skin_vst_dist_hm ~ health/subject, data=as(sample_data(skin_vst),"data.frame"))
#perm_vst_hm
```
Differential abundance analysis using DESeq2
This analysis uses raw counts... does not do a vst on the data, but does other data transformation to account for differences in library size and to stabilise variance within groups.
```{r, echo=FALSE}
#Subset raw count data to skin only
mydata_sub_skin<-subset_samples(mydata_sub, sampletype1%in%c("control_skin","diabetic_skin"))
#Filter OTUs seen less than 3 times in less than 20% of samples
mydata_sub_skin_filter<-filter_taxa(mydata_sub_skin, function(x) sum(x > 3) > (0.2*length(x)), TRUE)
#convert to DESeq2 object and specify experimental design
skin_diff_dds<-phyloseq_to_deseq2(mydata_sub_skin_filter, ~ health + health:subject2, ignoreRank=TRUE)
#manually creating the coldata file to manually create the model matrix
write.csv(sample_data(mydata_sub_skin_filter),file="r_data/coldata.csv")
coldata <- read.csv("r_data/coldata.csv")
head(coldata)
#The specified model creates issues with some columns of the model matrix having only 0.  Creating the model matrix manually and removing the columns with only 0 as specified in the DESeq2 vignette 3.12.2
m1 <- model.matrix(~ health + subject2:health, coldata)
colnames(m1)
#unname(m1)
m1 <- m1[,-18]
#unname(m1)
m1 <- m1[,-15]
#unname(m1)
#Make sure log2 abundance differences are calculated compared to the control
skin_diff_dds$health <- relevel(skin_diff_dds$health, "control")
#Do diferential abundance analysis on diabetic vs control skin adding the custom matrix generated above using the argument "full".
skin_diff_dds<-DESeq(skin_diff_dds, full=m1, betaPrior=FALSE)
#Get results, order by p value, then create a matrix with only sig p value results and merge with the names of the taxa for each OTU.
res_skin <- results(skin_diff_dds, alpha=0.05, lfcThreshold = 1)
summary(res_skin, alpha=0.05)
sig_num<-sum(res_skin$padj < 0.05, na.rm=TRUE)
resOrdered_skin <- res_skin[order(res_skin$padj),]
skin_sig <- resOrdered_skin[1:sig_num,]
df_skin_sig<-data.frame(skin_sig)
res_Ordered_skin_rownames <- row.names(skin_sig)
skin_res_names <- tax_table(mydata_sub_skin)[res_Ordered_skin_rownames,]
skin_diff_merge <-merge(skin_res_names, skin_sig, by="row.names")
skin_diff_merge<-skin_diff_merge[order(skin_diff_merge$padj),]
skin_diff_merge[1:10,]
#Make the column Row.Names the actual row names
row.names(skin_diff_merge)<- skin_diff_merge[,1]
skin_diff_merge[,1]<-NULL
#Order sig results by log2FoldChange
skin_diff_merge_fco <- skin_diff_merge[order(-skin_diff_merge$log2FoldChange),]
skin_diff_merge_fco[1:10,]
```
Plot diff abund OTU counts for top 5 log2FC
```{r, echo=FALSE}
#make health a factor for plotting
health<- mapping$health
log2FC_1<-plotCounts(skin_diff_dds, gene=row.names(skin_diff_merge_fco)[1], intgroup=c("health","subject2"), returnData=TRUE)
#alternative ggplot way of plotting the data is commented out below
#log2FC_1$subject2health <- interaction(log2FC_1$subject2, log2FC_1$health)
#ggplot(aes(y = count, x = subject2health), data = log2FC_1) + geom_boxplot() #+ scale_y_log10()
boxplot(log2FC_1$count ~ log2FC_1$subject2*log2FC_1$health, log="y", las=2, main=c(row.names(skin_diff_merge_fco)[1]," log2fc=",skin_diff_merge_fco$log2FoldChange[1]))
log2FC_2<-plotCounts(skin_diff_dds, gene=row.names(skin_diff_merge_fco)[2], intgroup=c("health","subject2"), returnData=TRUE)
boxplot(log2FC_2$count ~ log2FC_2$subject2*log2FC_2$health, log="y", las=2, main=c(row.names(skin_diff_merge_fco)[2]," log2fc=",skin_diff_merge_fco$log2FoldChange[2]))
log2FC_3<-plotCounts(skin_diff_dds, gene=row.names(skin_diff_merge_fco)[3], intgroup=c("health","subject2"), returnData=TRUE)
boxplot(log2FC_3$count ~ log2FC_3$subject2*log2FC_3$health, log="y", las=2, main=c(row.names(skin_diff_merge_fco)[3]," log2fc=",skin_diff_merge_fco$log2FoldChange[3]))
log2FC_4<-plotCounts(skin_diff_dds, gene=row.names(skin_diff_merge_fco)[4], intgroup=c("health","subject2"), returnData=TRUE)
boxplot(log2FC_4$count ~ log2FC_4$subject2*log2FC_4$health, log="y", las=2, main=c(row.names(skin_diff_merge_fco)[4]," log2fc=",skin_diff_merge_fco$log2FoldChange[4]))
log2FC_5<-plotCounts(skin_diff_dds, gene=row.names(skin_diff_merge_fco)[5], intgroup=c("health","subject2"), returnData=TRUE)
boxplot(log2FC_5$count ~ log2FC_5$subject2*log2FC_5$health, log="y", las=2, main=c(row.names(skin_diff_merge_fco)[5]," log2fc=",skin_diff_merge_fco$log2FoldChange[5]))
log2FC_115<-plotCounts(skin_diff_dds, gene=row.names(skin_diff_merge_fco)[115], intgroup=c("health","subject2"), returnData=TRUE)
boxplot(log2FC_115$count ~ log2FC_115$subject2*log2FC_115$health, log="y", las=2, main=c(row.names(skin_diff_merge_fco)[115]," log2fc=",skin_diff_merge_fco$log2FoldChange[115]))
```
Plot counts for top 5 by p value
```{r, echo=FALSE}
resOrdered_skin <- res_skin[order(res_skin$padj),]
padj_1<-plotCounts(skin_diff_dds, gene=row.names(resOrdered_skin)[1], intgroup=c("health","subject2"), returnData=TRUE)
boxplot(padj_1$count ~ padj_1$subject2*padj_1$health, log="y", las=2, main=c(row.names(resOrdered_skin)[1]," padj= ", resOrdered_skin$padj[1]))
padj_2<-plotCounts(skin_diff_dds, gene=row.names(resOrdered_skin)[2], intgroup=c("health","subject2"), returnData=TRUE)
boxplot(padj_2$count ~ padj_2$subject2*padj_2$health, log="y", las=2, main=c(row.names(resOrdered_skin)[2]," padj= ", resOrdered_skin$padj[2]))
padj_3<-plotCounts(skin_diff_dds, gene=row.names(resOrdered_skin)[3], intgroup=c("health","subject2"), returnData=TRUE)
boxplot(padj_3$count ~ padj_3$subject2*padj_3$health, log="y", las=2, main=c(row.names(resOrdered_skin)[3]," padj= ", resOrdered_skin$padj[3]))
padj_4<-plotCounts(skin_diff_dds, gene=row.names(resOrdered_skin)[4], intgroup=c("health","subject2"), returnData=TRUE)
boxplot(padj_4$count ~ padj_4$subject2*padj_4$health, log="y", las=2, main=c(row.names(resOrdered_skin)[4]," padj= ", resOrdered_skin$padj[4]))
padj_5<-plotCounts(skin_diff_dds, gene=row.names(resOrdered_skin)[5], intgroup=c("health","subject2"), returnData=TRUE)
boxplot(padj_5$count ~ padj_5$subject2*padj_5$health, log="y", las=2, main=c(row.names(resOrdered_skin)[5]," padj= ", resOrdered_skin$padj[5]))
```
Get relative abundances for OTUs per group for the sig diff OTUs
```{r, echo=FALSE}
merge_skin<-merge_samples(mydata_sub_skin, "health")
#transform sample counts to %
mydata_sub_skin_relab<-transform_sample_counts(mydata_sub_skin, function(x) 100 * x/sum(x))
#Get vector of OTU names with p<0.05 in diff abundance analysis
OTUnames<-row.names(skin_sig)
#Get average abundance of each OTU across all skin samples
allsums<-(taxa_sums(mydata_sub_skin_relab)[OTUnames])
averagesums<-(taxa_sums(mydata_sub_skin_relab)[OTUnames])/nsamples(mydata_sub_skin)
allsums
averagesums
fivenum(allsums)
fivenum(averagesums)
#create vectors with names of control samples and diabetic samples for skin
control_skin_samples<-rownames(sample_data(mydata_sub_skin)[sample_data(mydata_sub_skin)[,"health"]=="control"])
diabetic_skin_samples<-rownames(sample_data(mydata_sub_skin)[sample_data(mydata_sub_skin)[,"health"]=="diabetes"])
#apply the mean & stdev functions to sig diff OTUs only, and for only control or skin samples
control_sig_otus_means<-as.matrix(apply(otu_table(mydata_sub_skin_relab)[OTUnames,control_skin_samples],1,FUN=mean))
control_sig_otus_sd<-as.matrix(apply(otu_table(mydata_sub_skin_relab)[OTUnames,control_skin_samples],1,FUN=sd))
diabetic_sig_otus_means<-as.matrix(apply(otu_table(mydata_sub_skin_relab)[OTUnames,diabetic_skin_samples],1,FUN=mean))
diabetic_sig_otus_sd<-as.matrix(apply(otu_table(mydata_sub_skin_relab)[OTUnames,diabetic_skin_samples],1,FUN=sd))
#Create a table of sig dif OTU names, taxa, adjp values, and average relative abundance for both eeach of control and diabetic skin groups.
#First make sure sig otu means values are in the same order as the results of sig dif test
control_sig_otus_means<-as.matrix(control_sig_otus_means[(rownames(skin_diff_merge)),])
control_sig_otus_sd<-as.matrix(control_sig_otus_sd[(rownames(skin_diff_merge)),])
diabetic_sig_otus_means<-as.matrix(diabetic_sig_otus_means[(rownames(skin_diff_merge)),])
diabetic_sig_otus_sd<-as.matrix(diabetic_sig_otus_sd[(rownames(skin_diff_merge)),])
#Now bind average abundances and sd to the results table
sig_diff_otus_results<-cbind(skin_diff_merge,control_sig_otus_means,control_sig_otus_sd, diabetic_sig_otus_means, diabetic_sig_otus_sd)
sig_diff_otus_results<-sig_diff_otus_results[order(-sig_diff_otus_results$control_sig_otus_means),]
#Make a table of all OTUs from skin comparing the average abundance and stdev for control and diabetic skin
control_skin_samples<-rownames(sample_data(mydata_sub_skin)[sample_data(mydata_sub_skin)[,"health"]=="control"])
diabetic_skin_samples<-rownames(sample_data(mydata_sub_skin)[sample_data(mydata_sub_skin)[,"health"]=="diabetes"])
control_skin_otus_means<-as.matrix(apply(otu_table(mydata_sub_skin_relab)[,control_skin_samples],1,FUN=mean))
control_skin_otus_sd<-as.matrix(apply(otu_table(mydata_sub_skin_relab)[,control_skin_samples],1,FUN=sd))
diab_skin_otus_means<-as.matrix(apply(otu_table(mydata_sub_skin_relab)[,diabetic_skin_samples],1,FUN=mean))
diabet_skin_otus_sd<-as.matrix(apply(otu_table(mydata_sub_skin_relab)[,diabetic_skin_samples],1,FUN=sd))
skin_otus_all<-cbind(control_av=control_skin_otus_means,control_sd=control_skin_otus_sd, diab_av=diab_skin_otus_means, diab_sd=diabet_skin_otus_sd)
colnames(skin_otus_all)<-c("cnt_av","cnt_sd","diab_av","diab_sd")
write.csv(skin_otus_all, file="skin_abund.csv")
```
Plots of Relative abundance of OTUs of interest per subject
```{r}
OTU_13<-data.frame(otu_table(rare_skin)["OTU_13",])
OTU_13<-t(OTU_13)
OTU_13<-OTU_13[order(rownames(OTU_13)),,drop=FALSE]
subject<-sample_data(rare_skin)[order(rownames(sample_data(rare_skin))),"subject"]
OTU_13<-cbind(OTU_13,subject)
OTU_13$"OTU_13"<-(OTU_13$"OTU_13"/30000)*100
ggplot(OTU_13, aes(x=subject, y=OTU_13)) + 
  geom_boxplot()+ylim(0,5)+labs(x = "Subject", y="Relative abundance (%)")+
  ggtitle("Kocuria palustris relative abundance by subject")
OTU_21<-data.frame(otu_table(rare_skin)["OTU_21",])
OTU_21<-t(OTU_21)
OTU_21<-OTU_21[order(rownames(OTU_21)),,drop=FALSE]
subject<-sample_data(rare_skin)[order(rownames(sample_data(rare_skin))),"subject"]
OTU_21<-cbind(OTU_21,subject)
OTU_21$"OTU_21"<-(OTU_21$"OTU_21"/30000)*100
ggplot(OTU_21, aes(x=subject, y=OTU_21)) + 
  geom_boxplot()+ylim(0,5)+labs(x = "Subject", y="Relative abundance (%)")+
  ggtitle("OTU21 relative abundance by subject")
OTU_2855257<-data.frame(otu_table(rare_skin)["2855257",])
OTU_2855257<-t(OTU_2855257)
OTU_2855257<-OTU_2855257[order(rownames(OTU_2855257)),,drop=FALSE]
OTU_2855257<-cbind(OTU_2855257,subject)
OTU_2855257$"2855257"<-(OTU_2855257$"2855257"/30000)*100
ggplot(OTU_2855257, aes(x=subject, y=OTU_2855257$"2855257")) + 
  geom_boxplot()+ylim(0,0.25)+labs(x = "Subject", y="Relative abundance (%)")+ggtitle("Modestobacter relative abundance by subject")
OTU_205559<-data.frame(otu_table(rare_skin)["205559",])
OTU_205559<-t(OTU_205559)
OTU_205559<-OTU_205559[order(rownames(OTU_205559)),,drop=FALSE]
OTU_205559<-cbind(OTU_205559,subject)
OTU_205559$"205559"<-(OTU_205559$"205559"/30000)*100
ggplot(OTU_205559, aes(x=subject, y=OTU_205559$"205559")) + 
  geom_boxplot()+ylim(0,0.055)+labs(x = "Subject", y="Relative abundance (%)")+ggtitle("Microbacteriaceae relative abundance by subject")
#Highest by GINI index
OTU_16101<-data.frame(otu_table(rare_skin)["16101",])
OTU_16101<-t(OTU_16101)
OTU_16101<-OTU_16101[order(rownames(OTU_16101)),,drop=FALSE]
OTU_16101<-cbind(OTU_16101,subject)
OTU_16101$"16101"<-(OTU_16101$"16101"/30000)*100
ggplot(OTU_16101, aes(x=subject, y=OTU_16101$"16101")) + 
  geom_boxplot()+ylim(0,0.15)+labs(x = "Subject", y="Relative abundance (%)")+ggtitle("Caldicellulosiruptor saccharolyticus relative abundance by subject")
```
#Now for random forrest - can I paste Sunaina's script here? Let's see....
##Loading packages and data files in phyloseq
```{r, echo=FALSE}
list.of.packages <- c("ggplot2", "Rcpp", "phyloseq","cluster","randomForest","caret","e1071","vegan","scales","grid","doParallel","foreach")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
library(phyloseq)
library(cluster)
library(ggplot2)
library(randomForest)
library(caret)
library(e1071)
library(vegan)
library("scales")
library("grid")
library("doParallel")
library("foreach")
#rich_biom = ("r_data/otu_table_mc2_w_tax_no_pynast_failures_json.biom")
#mappingfile = "r_data/mapping_diabetes_all.txt"
#treefile = ("r_data/otus_cf_aligned.tre")
#mydata = import_biom(rich_biom,treefile)
#map = import_qiime_sample_data(mappingfile)
#mergedata = merge_phyloseq(mydata,map)
#mydataedgeR = import_biom(rich_biom,treefile,parseFunction=parse_taxonomy_greengenes)
#edgerdata= merge_phyloseq(mydataedgeR,map)
#tax_table(edgerdata) <- tax_table(edgerdata)[, -8]
```
###Basic Random Forest
```{r}
wh0 = genefilter_sample(rare_skin, filterfun_sample(function(x) x > 0), A = 0.10 * nsamples(rare_skin))
rare_skin_filter <- prune_taxa(wh0, rare_skin)
```{r,echo=FALSE}
t_otutable<-t(otu_table(rare_skin_filter))
Health <- as.factor(sample_data(rare_skin_filter)$health)
patientname<-sample_data(rare_skin_filter)$subject
rf_data<-data.frame(Health,patientname,t_otutable)
set.seed(250)
rf_data$Health<-ordered(rf_data$Health,levels=c("diabetes","control"))
uniqID<-length(unique(rf_data$patientname))
trainingID<-sample(unique(rf_data$patientname),round(uniqID*0.5),replace=FALSE)
train<-rf_data[rf_data$patientname %in% trainingID,]
test<-rf_data[!rf_data$patientname %in% trainingID,]
train<-train[-2]
test<-test[-2]
mtry <- tuneRF(train[-1],train$Health, ntreeTry=200,stepFactor=1.5,improve=0.01, trace=TRUE, plot=TRUE)
best.m <- mtry[mtry[, 2] == min(mtry[, 2]), 1]
print(mtry)
print(best.m)
rf <-randomForest(Health~.,data=train, mtry=best.m, importance=TRUE,ntree=200)
predictionsrandomF <- predict(rf,test[-1])
confusionMatrix(test$Health,predictionsrandomF) 
```

###Naive Bayes
```{r,echo=FALSE}
model <- naiveBayes(Health ~ ., train)
predBayes<-predict(model, test[-1])
confusionMatrix(test$Health,predBayes)
```

###rpart
```{r,echo=FALSE}
fitControl = trainControl( method = "cv", number = 10 )
rparttrain <- train(Health ~ ., data = train, method = "rpart", trControl = fitControl)
predrpart <- predict(rparttrain, test[-1])
confusionMatrix(test$Health, predrpart)
```

###Finding important predictors for RF using Gini importance
```{r,echo=FALSE}
predictors<-importance(rf)
varImpPlot(rf)
predictors<-data.frame(names=rownames(predictors),predictors)
predictors<-predictors[order(-predictors[,5]),]
predictors<-predictors[predictors$MeanDecreaseGini>0.05,]
otunames<- predictors$names
otunames<-as.character(otunames)
otunames<-gsub("^X","",otunames)
subsetOTU<-prune_taxa(otunames,rare_skin_filter)
otulist<-data.frame(tax_table(subsetOTU))[otunames,]
predictors<-cbind(otulist,predictors)
##Compare results to DESeq 2 sig diff results - merge dataframes?
rownames(predictors)<-otunames
compare<-merge(predictors, skin_diff_merge,by=0,all=TRUE)
#Add abundances from skin
abund_sig_merge<-skin_otus_all[compare[,1],]
```
Finding average and sd abundance in skin of OTUs of interest
```{r}
#Define list of OTUs of interest - from a spreadsheet comparing results from Kruskal and Wallice tst for significant differences, and Gini Index
sig_otus_interest<-c("274011","818388","2855257","New.CleanUp.ReferenceOTU10998","948180","205559","1002005","1048194","993474","569992","New.ReferenceOTU181","403359","4418684","New.ReferenceOTU153","688714","952388","1093944","369436","552687","1108726","349901","16101")
abund_sig_otus_interest<-average_abund_sd_skin[sig_otus_interest,]
tax_sig_otus_interest<-tax_table(rare_mydata_sub)[sig_otus_interest,]
abund_sig_otus_interest<-as.matrix(cbind(abund_sig_otus_interest,tax_sig_otus_interest))
abund_sig_otus_interest
#Now get abundance for GINI indext sig results.
gini_otus_abund<-average_abund_sd_skin[otunames,]
tax_gini_otus<-tax_table(rare_mydata_sub)[otunames,]
gini_otus_abund<-cbind(gini_otus_abund,tax_gini_otus)
```
DESeq 2 analysis for differential abundance between diabetic skin and wounds
```{r}
#subset to diabetic skin and wounds only
mydata_diabetes<-subset_samples(mydata_sub, health=="diabetes")
mydata_diabetes
#Filter OTUs seen less than 3 times in less than 20% of samples
mydata_diabetes_sub<-filter_taxa(mydata_diabetes, function(x) sum(x > 3) > (0.2*length(x)), TRUE)
#Convert to DESeq2 object and specify the experimental design - don't use numbers for factors!
diabetes_dds<-phyloseq_to_deseq2(mydata_diabetes_sub, ~ sampletype1 + sampletype1:subject)
#Make sure log2 abundance differences are calculated compared to skin
diabetes_dds$sampletype1 <- relevel(diabetes_dds$sampletype1, "diabetic_skin")
diabetes_dds<-DESeq(diabetes_dds)
#Get results and order by adjusted p value
res_diabetes<-results(diabetes_dds,alpha=0.05, lfcThreshold = 1)
summary(res_skin, alpha=0.05)
sig_num_diabetes<-sum(res_diabetes$padj < 0.05, na.rm=TRUE)
resOrdered_diabetes <- res_diabetes[order(res_diabetes$padj),]
sig_diab <- resOrdered_diabetes[1:sig_num_diabetes,]
df_sig_diab<-data.frame(sig_diab)
res_Ordered_diabetes_rownames <- row.names(sig_diab)
res_diabetes_names <- tax_table(mydata_diabetes_sub)[res_Ordered_diabetes_rownames,]
diabetes_diff_merge <-merge(res_diabetes_names, sig_diab, by="row.names")
diabetes_diff_merge<-diabetes_diff_merge[order(diabetes_diff_merge$padj),]
diabetes_diff_merge[1:10,]
#Make the column Row.Names the actual row names
row.names(diabetes_diff_merge)<- diabetes_diff_merge[,1]
diabetes_diff_merge[,1]<-NULL
write.table(diabetes_diff_merge, "r_data/diabetes_diff_merge.txt", sep="\t"")
#Order sig results by log2FoldChange
diabetes_diff_merge_fco <- diabetes_diff_merge[order(-diabetes_diff_merge$log2FoldChange),]
diabetes_diff_merge_fco[1:10,]
#Plot top 5 by log fold change
#Make a factor for plotting
sampletype1<- sample_data(mydata_diabetes_sub)$sampletype1
log2FC_1<-plotCounts(diabetes_dds, gene=row.names(diabetes_diff_merge_fco)[1], intgroup=c("sampletype1","subject2"), returnData=TRUE)
boxplot(log2FC_1$count ~ log2FC_1$subject2*log2FC_1$sampletype1, log="y", las=2, main=c(row.names(diabetes_diff_merge_fco)[1]," log2fc=",diabetes_diff_merge_fco$log2FoldChange[1]))
```
Get relative abundances for OTUs per group for the sig diff OTUs
```{r, echo=FALSE}
#transform sample counts to %
mydata_diabetes_sub_relab<-transform_sample_counts(mydata_diabetes_sub, function(x) 100 * x/sum(x))
#Get vector of OTU names with p<0.05 in diff abundance analysis
OTUnames_diab_diff<-row.names(diabetes_diff_merge)
#create vectors with names of control samples and diabetic samples for skin
diabetic_wound_samples<-rownames(sample_data(mydata_diabetes_sub)[sample_data(mydata_diabetes_sub)[,"sampletype1"]=="wound"])
diabetic_skin_samples2<-rownames(sample_data(mydata_diabetes_sub)[sample_data(mydata_diabetes_sub)[,"health"]=="diabetes"])
#apply the mean & stdev functions to sig diff OTUs only, and for only wound or skin samples
diab_skin_sig_otus_means<-as.matrix(apply(otu_table(mydata_diabetes_sub_relab)[OTUnames_diab_diff,diabetic_skin_samples2],1,FUN=mean))
wound_sig_otus_means<-as.matrix(apply(otu_table(mydata_diabetes_sub_relab)[OTUnames_diab_diff,diabetic_wound_samples],1,FUN=mean))
diab_skin_sig_otus_sd<-as.matrix(apply(otu_table(mydata_diabetes_sub_relab)[OTUnames_diab_diff,diabetic_skin_samples2],1,FUN=sd))
wound_sig_otus_sd<-as.matrix(apply(otu_table(mydata_diabetes_sub_relab)[OTUnames_diab_diff,diabetic_wound_samples],1,FUN=sd))
#Create a table of sig dif OTU names, taxa, adjp values, and average relative abundance for both each of wound and diabetic skin groups.
#First make sure sig otu means values are in the same order as the results of sig dif test
diab_skin_sig_otus_means<-as.matrix(diab_skin_sig_otus_means[(rownames(diabetes_diff_merge)),])
diab_skin_sig_otus_sd<-as.matrix(diab_skin_sig_otus_sd[(rownames(diabetes_diff_merge)),])
wound_sig_otus_means<-as.matrix(wound_sig_otus_means[(rownames(diabetes_diff_merge)),])
wound_sig_otus_sd<-as.matrix(wound_sig_otus_sd[(rownames(diabetes_diff_merge)),])
#Now bind average abundances and sd to the results table
sig_diff_diab_otus_results<-cbind(diabetes_diff_merge,diab_skin_sig_otus_means,diab_skin_sig_otus_sd, wound_sig_otus_means, wound_sig_otus_sd)
#Order by decreasing abundance in wounds
sig_diff_diab_otus_results<-sig_diff_diab_otus_results[order(-sig_diff_diab_otus_results$wound_sig_otus_means),]
```
Plots of Relative abundance of OTUs of interest per subject
```{r}
#Make rarefied table for diabetic samples
rare_mydata_diabetes_sub<-rarefy_even_depth(mydata_diabetes_sub, sample.size=30000, trimOTUs=TRUE, rngseed=552)
subject<-sample_data(rare_mydata_diabetes_sub)[order(rownames(sample_data(rare_mydata_diabetes_sub))),"subject"]
sampletype1_diab<-sample_data(rare_mydata_diabetes_sub)[order(rownames(sample_data(rare_mydata_diabetes_sub))),"sampletype1"]
#Plot boxplots of relative abundance from rarefied counts for sig diff OTUs from diabetes_diff_merge grouped by patient and separated into wound and skin sample types
for (wound_otu in row.names(diabetes_diff_merge)){
  wound_otu_fig<-t(data.frame(otu_table(rare_mydata_diabetes_sub)[wound_otu,]))
  wound_otu_fig<-wound_otu_fig[order(rownames(wound_otu_fig)),,drop=FALSE]
  wound_otu_fig<-cbind(wound_otu_fig,subject,sampletype1_diab)
  wound_otu_fig[1]<-(wound_otu_fig[,1]/30000)*100
  wound_otu_fig$subjectsampletype1<-interaction(wound_otu_fig$subject, wound_otu_fig$sampletype1)
  print(ggplot(wound_otu_fig, aes(x=subject, y=wound_otu_fig[,1], fill= sampletype1, dodge=subject)) + 
  geom_boxplot()+labs(x = "Subject", y="Relative abundance (%)")+
  ggtitle(wound_otu))
}
```
Distances within groups for diabetic skin (contra) and control skin


Generate boxplots based on weighted unifrac distances between diabetic adjacent and contrateral skin, and wound swabs and debridement
```{r, echo=FALSE, fig.width=10, fig.height=5}
dist_skin<-read.table("r_data/control_diab_box.txt", header=TRUE, sep="\t")
boxplot(dist_skin, las=2, main="Unifrac distances between adjacent or contralateral\n diabetic skin and wounds", ylab="Weighted unifrac distance", cex.lab=0.75, cex.main=0.75, cex.axis=0.75, ylim=c(0,0.8))

